Title TBD

### 2019-07-02 main/draft.tex
New version

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
\title{Kids First or Marriage First?}
\author{Egor Kozlov}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
\maketitle

\begin{abstract}
Unplanned pregnancies often trigger marriage for unmarried couples. I document two robust empirical facts. First, women who have their first child before or at the year of marriage are statistically more likely to be divorced in the future than those who marry before they have a child. Second, this difference is more pronounced for women with higher income and educational attainment, although it is less common for them to have their kids before marriage. I estimate a lifecycle model with dynamic bargaining, marriage selection and unplanned pregnancy risks. I show that the first fact can be attributed to household bargaining friction, as women agree to enter lower quality marriages to avoid being a single mother, and  the second fact is consistent with childcare requiring both time and money, as this suggests that the costs of being a single mother are higher for more productive women. Qualitatively, this sheds some light on household formation and dissolution and importance of accounting for heterogeneous time costs. Qualitatively, the model suggests sizable causal effect of unplanned pregnancies on observable divorce rates.
\end{abstract}

\section{Introduction}
As of 2018, [40\%] of new births occurred by unmarried mothers. This number is rising for decades, but, more importantly, it is rather heterogeneous, reaching three-quarters for mothers who are African American, [...] under 25 and [...] for high school or less in education. But even for educated and older women this share is substantial [...]. An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not commong: according to National Survey of Family And Growth, the median duration of cohabitation is 22 months, and median duration of cohabitation if it ends by marriage is 21.\footnote{Source: Copen et al, \cite{copen}}. Unmarried unions with children are still pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents (as opposed to 21\% with single mothers).

These numbers have important economic implications: children born by unmarried women, as documented by, for instance Kearney, Levine, 2017, display significantly worse future outcomes in terms of education and income, and this pattern survives controlling for many dimensions. Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.

This pairs well with common perception that some couples marry at or shortly after the moment they make a baby. If cohabitation is viewed as uncommitted relationship, unexpected fertility may serve as a trigger to obtain more commitment. Narrowly, we can interpret this as shotgun marriage story, however in the times with large abortion access many unplanned pregnancies are aborted, and cohabitating couples seem to \emph{choose} to keep an unplanned baby. The whole role of marriage changes as it is more socially acceptable to cohabitate, though it is much more conventional to do it before having kids. As highlighted by, for instance, Lafortune, 2018, despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.

This paper highlights one important aspect of this: I argue that when marriage decision is driven by having an unexpected child, bargaining friction makes marital outcome worse. The source of this friction is the large childcare costs and unequal relative exposure to them: as women bear relatively more childcare costs, they are more likely to enter relatively lower quality marriages as a commitment to share a part of these costs. This can rationalize observable differences in marital outcomes between couples with and without out-of-wedlock births.

I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups. I then provide a model incorporating the bargaining mechanism that can rationalize these differences. I estimate the model parameters to match observable patterns in the data and show what it implies about the role of unequal exposure to childcare costs in observable marriage selection and divorce patterns. Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by [...], especially in high-income groups.

>>>>

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
\title{Title TBD}
\author{Egor Kozlov}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
\maketitle

\begin{abstract}
[WIP] Many couples marry shortly after woman gets pregnant, that is often referred as a shotgun marriage. I show that in American Community Survey data couples who experienced the shotgun marriage divorce more often in the future. This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of shotgun marriages is lower among them. I argue the difference in divorce rates can be plausibly explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality, and in the data having kids before marriage is a plausible proxy for having lower-quality marriage. I argue that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs. I estimate a structural model endogenizing marriage selection and use it to argue that marriage quality, as opposed to different composition of the groups, drives the major part of the observable difference in divorce probabilities.
\end{abstract}

\section{Introduction}

Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple. This practice does not solely refer back to more conservative times: in the modern US data  around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014 sample}. And, although it seems that abortion and contraception access would make give couples more choice, in the data such couples still look like something forced them to marry each other. In particular, their marriages appear less stable: despite of being less than quarter of the population such couples account for more than one-third of divorces\footnote{ACS 2009--2017}.

The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) if conventionally considered as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

In addition to child development consequences of unstable marriages, divorces have aggregate consequences on their own. For instance, they appear as important driver of labor market choices and outcomes of women. Additionally, being divorced with young children is considered as lowering re-marriage prospective, and this can affect total fertility rates. Finally, the question of marriage selection and modeling divorces is curious by its own from prospectives of economic theory: if shotgun marriage predicts lower marriage quality and higher likelihood of divorce, we can use it as identifying variation for studying intra-household questions, that is relevant for developing realistic household models and understanding things that are not easy to observe directly.

My main empirical contribution is describing association between having shotgun marriages and having a divorce. Although it is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations. In particular, [...]. The heterogeneity of this pattern can be informative of the channels and serves as a setup for the next model part of the paper.

As additional findings, I establish few other results. [... ... ...]

To deal with obvious endogeneity and selection affecting the comparisons above I set up a structural model that is capable of understanding of anatomy of the result observed in the data. The model is based on growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs. The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them. The key driving shock is that right before the moment people decide on their marriage they may unexpectedly get pregnant, and combination of it with differential costs of childcare generates desirable patterns in marriage selection. The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data extremely well.

The key conclusion of the structural part is that in estimated model couples who experienced an unplanned pregnancy shock agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data. he model delivers various decomposition, splitting observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality. As it turns out, the last factor is the most important contributor to the difference, and its contribution is [off the similar magnitude than the observable difference]. Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e. who complied with the shock), although there is some additional distress of having a baby on those who would have married anyway.

Additionally, the model delivers few nontrivial counterfactual predictions. First, in the model major part of divorces that happen early in life are driven by unexpected pregnancies, though divorces happening later are not. Second, general impact of pressure of becoming single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...]. [Lastly, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a good proxy for an inferior marriage quality, and therefore the difference between two groups is not purely based on demography and economic outcomes. Therefore observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children.

The causal lesson is that treat of being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be social rather than economic issue.
